---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

```{r echo=FALSE,fig.width=3.5,out.extra='style="float: right;"',fig.cap='Tracklines derived from all Argos and GPS location estimates',warning=FALSE,message=FALSE}
library(kotzeb0912)
library(dplyr)
library(ggmap)

data("kotzeb0912_locs")

locs <- kotzeb0912_locs %>% 
  dplyr::select(deployid,date_time,latitude,longitude) %>% 
  dplyr::filter(!is.na(latitude)) %>% 
  dplyr::arrange(deployid,date_time) %>% 
  data.frame()

med_lng <- median(locs$longitude)
med_lat <- median(locs$latitude) -3

map <- get_map(location=c(med_lng,med_lat),
                     zoom=4,source="google",maptype='terrain',color="bw")
gg <- ggmap(map) + geom_point(data=locs,mapping=aes(x=longitude,y=latitude),
                              alpha=0.25,size=0.75)
gg
```

Bearded seals (_Erignathus barbatus_) are one of the most important subsistence
resources for the indigenous people of coastal northern and western Alaska, as
well as key components of Arctic marine ecosystems, yet relatively little about
their abundance, seasonal distribution, migrations, or foraging behaviors has
been documented scientifically. Ice-associated seal populations may be
negatively impacted by offshore oil and gas development as well as by climate
change. Our ability to predict impacts, however, is limited by inadequate
knowledge of seal population structure and foraging ecology. By working
cooperatively with Alaska Native subsistence hunters we developed methods for
live-capturing bearded seals in the Chukchi Sea using nets set in the shallow
coastal waters where bearded seals were foraging.  Capture efforts were based
out of Kotzebue and various locations in the North Slope Borough from Wainwright
to Barrow in June and July from 2009 to 2012.  In all, 7 seals were caught (2
adults and 5 sub-adults; 4 males and three females; ranging in length and weight
from 159 cm and 116 kg to 216 cm and 253 kg), all from Kotzebue Sound.  Each
seal was sampled for health and condition and released with three different types
of satellite-linked bio-loggers: the SPOT5, attached
to a rear flipper, provided information on the timing of hauling out and on the
seal’s location for up to three years; the MK10, glued to the top of a
seal’s head, provided Argos estimates of location and also provided
data on the timing and depths of dives, for up to ten months; the Mk10-AF, also
glued to the top of the head, provided GPS quality locations in addition to the
Argos estimates of location and dive behavior data.
