---
title: "Kotzebue Bearded Seal Telemetry Data 2009-2013"
authors:
- name: Josh M. London
  orcid: 0000-0002-3647-5046
  affiliation: National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA, Seattle, WA, USA
- name: Michael F. Cameron
  affiliation: National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA, Seattle, WA, USA
- name: Peter L. Boveng
  affiliation: National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA, Seattle, WA, USA
output: 
  html_document:
    theme: flatly
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
# opts_chunk$set()
```

```{r libs,include=FALSE}
library(kotzeb0912)
library(dplyr)
library(ggmap)
```
----------------------------
#### Josh M. London, Michael F. Cameron, and Peter L. Boveng
National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA, Seattle, WA, USA

Last updated: `r Sys.Date()`

## Executive Summary

```{r echo=FALSE,fig.width=5,out.extra='style="float: right;"',fig.cap='Tracklines derived from all Argos and GPS location estimates',warning=FALSE,message=FALSE}
data("kotzeb0912_locs")

locs <- kotzeb0912_locs %>% 
  dplyr::select(deployid,date_time,latitude,longitude) %>% 
  dplyr::filter(!is.na(latitude)) %>% 
  dplyr::arrange(deployid,date_time) %>% 
  data.frame()

med_lng <- median(locs$longitude)
med_lat <- median(locs$latitude) -3

map <- get_map(location=c(med_lng,med_lat),
                     zoom=4,source="google",maptype='terrain',color="bw")
gg <- ggmap(map) + geom_point(data=locs,mapping=aes(x=longitude,y=latitude),
                              alpha=0.25,size=0.75)
gg
```

Bearded seals (_Erignathus barbatus_) are one of the most important subsistence
resources for the indigenous people of coastal northern and western Alaska, as
well as key components of Arctic marine ecosystems, yet relatively little about
their abundance, seasonal distribution, migrations, or foraging behaviors has
been documented scientifically. Ice-associated seal populations may be
negatively impacted by offshore oil and gas development as well as by climate
change. Our ability to predict impacts, however, is limited by inadequate
knowledge of seal population structure and foraging ecology. By working
cooperatively with Alaska Native subsistence hunters we developed methods for
live-capturing bearded seals in the Chukchi Sea using nets set in the shallow
coastal waters where bearded seals were foraging.  Capture efforts were based
out of Kotzebue and various locations in the North Slope Borough from Wainwright
to Barrow in June and July from 2009 to 2012.  In all, 7 seals were caught (2
adults and 5 sub-adults; 4 males and three females; ranging in length and weight
from 159 cm and 116 kg to 216 cm and 253 kg), all from Kotzebue Sound.  Each
seal was sampled for health and condition and released with three different types
of satellite-linked bio-loggers: the SPOT5, attached
to a rear flipper, provided information on the timing of hauling out and on the
seal’s location for up to three years; the MK10, glued to the top of a
seal’s head, provided Argos estimates of location and also provided
data on the timing and depths of dives, for up to ten months; the Mk10-AF, also
glued to the top of the head, provided GPS quality locations in addition to the
Argos estimates of location and dive behavior data.

## Data Components

The **kotzeb0912** package is a data package for distribution of core data 
products resulting from this study. There are 7 data products distributed with
this package:

1. `kotzeb0912_locations`
    - Argos and GPS location estimates for all deployments
2. `kotzeb0912_status`
    - tag status and performance messages for all deployments
3. `kotzeb0912_divedepth`
    - dive-depth histogram data; cleaned and tidy'd
4. `kotzeb0912_diveduration`
    - dive-duration histogram data; cleaned and tidy'd
5. `kotzeb0912_tad`
    - time-at-depth histogram data; cleaned and tidy'd
6. `kotzeb0912_timelines`
    - hourly percent dry (haul-out) data; cleaned and tidy'd
7. `kotzeb0912_deployments`
    - bearded seal morphometric and tag deployment data


## Bearded Seals

Seven bearded seals were captured between 2009 and 2012.

## Telemetry Devices

All of the telemetry devices deployed in this study were manufactured by
Wildlife Computers (Redmond, Washington, USA) [^1]. All of the tags relied on
the Argos satellite network for location esimates and transfer of data. A few
tags were equipped with Fastloc-GPS capabilities that provided a limited
number of GPS quality locations [^2]. In addition to the Fastloc-GPS tags, 
flipper-mounted SPOT style tags were also deployed. These tags provided location
and haul-out data for longer duration than the Mk10 tags (attached to the hair
which molts each spring).

[^1]: mention of specific products or manufacturers does not constitute an 
endorsement by NMFS, NOAA, or the U.S. Department of Commerce

[^2]: fastloc-GPS was used sparingly because the technology was relatively new
at the time of this study and there is a significant increase in battery
consumption when using Fastloc-GPS compared to Argos.

## Data Gathering and Processing

### Tag Programming

There were three tag types/models deployed during this project

1. Mk10
2. Mk10-AF
3. SPOT5

#### Mk10 Programming Summary

#### Mk10-AF Programming Summary

#### SPOT5 Programming Summary

### Argos Least-Squared vs. Kalman Processing

In 2011, Argos changed their location estimation algorithm to include a Kalman
filter algorithm. This replaced their previous algorithm which relied on a 
least-squares process. The Kalman filter algorithm provides additional error
data which is critical for many movement models. When this change was made, 
all data for tags in this project were reprocessed back to January of 2008. In
most cases, for each location estimate in the least-squares dataset, there is a
corresponding Kalman filter estimate for that satellite pass.

The reprocessed data and the originally delivered least-squares data were
merged and reconciled by Wildlife Computers within their data portal. In this
process, locations were matched by _ptt_, _date-time_, _satellite_ and 
_pass duration_. _date-time_ and _pass duration_ were matched with fuzzy
logic (i.e. they were allowed to match exactly). In all cases, the reprocessed
kalman locations were determined to be authoritative.

### Wildlife Computers Data Portal

Data included in this package were downloaded from the Wildlife Computers
Data Portal using the **wcUtils** package.

## Publications and Data Availability

The final contract report for this research is available from the [Bureau of
Ocean and Energy Management](http://www.boem.gov/BOEM-Newsroom/Library/Publications/2013/BOEM_2013_001150_pdf.aspx)

This R data package and associated vignette documents should be cited as

> London, Josh M., M.F. Cameron, and P.L. Boveng. 2015. kotzeb0912: An R package
of data from a satellite telemetry study of bearded seals in Alaska.

## Acknowledgments and Funding

The research described here and the included data were obtained with significant
financial contributions from the U.S. Department of Interior's Bureau of Ocean
and Energy Management (BOEM) [^3] and the U.S. Department of Commerce's National
Oceanic and Atmospheric Administration.

[^3]: funding administered under the Inter-agency Agreement M07RG13317

