---
title: "kotzeb0912: An R package of data from a satellite telemetry study of bearded seals in Alaska"
authors:
- name: Josh M. London
  orcid: 0000-0002-3647-5046
  affiliation: National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA, Seattle, WA, USA
- name: Michael F. Cameron
  affiliation: National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA, Seattle, WA, USA
- name: Peter L. Boveng
  affiliation: National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA, Seattle, WA, USA
output: 
  rmarkdown::html_vignette:
    css: https://raw.githubusercontent.com/jasonm23/markdown-css-themes/gh-pages/markdown7.css
    fig_width: 4
    fig_height: 4
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
# opts_chunk$set()
```

```{r libs,include=FALSE}
library(kotzeb0912)
library(dplyr)
library(ggmap)
```

----------------------------

> **Josh M. London**  
Polar Ecosystems Program  
Alaska Fisheries Science Center, NOAA  
Seattle, Washington, USA  
orcid: 0000-0002-3647-5046  
josh.london@noaa.gov  
>
> **Michael F. Cameron**  
Polar Ecosystems Program  
Alaska Fisheries Science Center, NOAA  
Seattle, Washington, USA  
>
> **Peter L. Boveng**  
Polar Ecosystems Program  
Alaska Fisheries Science Center, NOAA  
Seattle, Washington, USA  

Last updated: `r Sys.Date()`

----------------------------


# Executive Summary

```{r echo=FALSE,fig.width=3.5,out.extra='style="float: right;"',fig.cap='Tracklines derived from all Argos and GPS location estimates',warning=FALSE,message=FALSE}
data("kotzeb0912_locs")

locs <- kotzeb0912_locs %>% 
  dplyr::select(deployid,date_time,latitude,longitude) %>% 
  dplyr::filter(!is.na(latitude)) %>% 
  dplyr::arrange(deployid,date_time) %>% 
  data.frame()

med_lng <- median(locs$longitude)
med_lat <- median(locs$latitude) -3

map <- get_map(location=c(med_lng,med_lat),
                     zoom=4,source="google",maptype='terrain',color="bw")
gg <- ggmap(map) + geom_point(data=locs,mapping=aes(x=longitude,y=latitude),
                              alpha=0.25,size=0.75)
gg
```

Bearded seals (_Erignathus barbatus_) are one of the most important subsistence
resources for the indigenous people of coastal northern and western Alaska, as
well as key components of Arctic marine ecosystems, yet relatively little about
their abundance, seasonal distribution, migrations, or foraging behaviors has
been documented scientifically. Ice-associated seal populations may be
negatively impacted by offshore oil and gas development as well as by climate
change. Our ability to predict impacts, however, is limited by inadequate
knowledge of seal population structure and foraging ecology. By working
cooperatively with Alaska Native subsistence hunters we developed methods for
live-capturing bearded seals in the Chukchi Sea using nets set in the shallow
coastal waters where bearded seals were foraging.  Capture efforts were based
out of Kotzebue and various locations in the North Slope Borough from Wainwright
to Barrow in June and July from 2009 to 2012.  In all, 7 seals were caught (2
adults and 5 sub-adults; 4 males and three females; ranging in length and weight
from 159 cm and 116 kg to 216 cm and 253 kg), all from Kotzebue Sound.  Each
seal was sampled for health and condition and released with three different types
of satellite-linked bio-loggers: the SPOT5, attached
to a rear flipper, provided information on the timing of hauling out and on the
seal’s location for up to three years; the MK10, glued to the top of a
seal’s head, provided Argos estimates of location and also provided
data on the timing and depths of dives, for up to ten months; the Mk10-AF, also
glued to the top of the head, provided GPS quality locations in addition to the
Argos estimates of location and dive behavior data.

----------------------------

# Bearded Seals

Seven bearded seals were captured between 2009 and 2012 and 14 tags were deployed.
Each animal was released with a head-mounted satellite tag and a flipper-mounted
satellite tag.

```{r deploy-table,echo=FALSE}
data('kotzeb0912_deployments')
d <- dplyr::arrange(kotzeb0912_deployments,speno,deployid) %>% 
  dplyr::select(-ptt)
hdrs <- c("Speno","Capture Time (UTC)","Age","Sex","DeployID","Tag Location")
knitr::kable(d,col.names = hdrs)
```

# Telemetry Devices

All of the telemetry devices deployed in this study were manufactured by
Wildlife Computers (Redmond, Washington, USA) [^2]. All of the tags relied on
the Argos satellite network for location esimates and transfer of data. A few
tags were equipped with Fastloc-GPS capabilities that provided a limited
number of GPS quality locations [^3]. In addition to the Fastloc-GPS tags, 
flipper-mounted SPOT style tags were also deployed. These tags provided location
and haul-out data for longer duration than the Mk10 tags (attached to the hair
which molts each spring).

[^2]: mention of specific products or manufacturers does not constitute an 
endorsement by NMFS, NOAA, or the U.S. Department of Commerce

[^3]: fastloc-GPS was used sparingly because the technology was relatively new
at the time of this study and there is a significant increase in battery
consumption when using Fastloc-GPS compared to Argos.

# Data Gathering and Processing

## Tag Programming

There were three tag types/models deployed during this project

1. Mk10
2. Mk10-AF
3. SPOT5

### Mk10 (-AF) Programming Summary

#### Histogram Data

    Histogram Data sampling interval
      ~ 10 seconds
    Dive Maximum Depth (m)
      ~ 14 bins:	10;30;50;70;90;100;150;200;250;300;400;500;600;>600
    Dive Duration (min)
      ~ 14 bins:	1;2;3;4;6;8;10;12;16;20;30;40;50;>50
    Time-at-Temperature (C)	
      ~ disabled
    Time-at-Depth (m)
      ~ 14 bins:	4;10;30;50;70;90;100;150;200;250;300;400;500;>500
    20-min time-line
      ~ disabled
    Hourly % time-line (low resolution)
      ~ enabled
    Hourly % time-line (high resolution)
      ~ disabled
    Light-level locations
      ~ disabled
  
#### Histogram Collection

    Hours of data summarized in each histogram
      ~ 6
    Histograms start at
      ~ GMT	03:00

#### Dive & Timeline Definition

    Depth reading to determine start and end of dive
      ~ Wet/Dry (Mk10)
      ~ 2m (Mk20-A)
    Ignore dives shallower than	
      ~ 4m
    Ignore dives shorter than
      ~ 1 min
    Depth threshold for timelines
      ~ Wet/Dry
    Haulout Definition
      ~ A minute is "dry" if Wet/Dry sensor is dry for any **30** seconds in a minute
      ~ Enter haulout state after **5** consecutive dry minutes
      ~ Exit haulout state if wet for any **50** seconds in a minute
  
#### Argos Transmissions
  
    Transmission Hours
      ~ 0-23
    Transmission Days
      ~ All Days
    Transmission Monhts
      ~ All Months
    Daily Transmission Cap
      ~ 250 Transmissions

#### Fast-GPS Settings

    Fast-GPS sampling interval
      ~ 360 minutes
    Deployment Latitude
      ~ 66.75 degrees
    Deployment Longitude
      ~ -163 degrees
    Deployment Altitude
      ~ 0 m
    Transmit hours
      ~ 0 - 23
    Fast-GPS Collection Days
      ~ January	4, 8, 12, 16, 20, 24, 28
      ~ February	1, 5, 9, 13, 17, 21, 25
      ~ March	1, 5, 9, 13, 17, 21, 25, 29
      ~ April	2, 6, 10, 14, 18, 22, 26, 30
      ~ May	4, 8, 12, 16, 20, 24, 28
      ~ June	1, 5, 9, 10, 14, 18, 22, 26, 30
      ~ July	4, 8, 12, 16, 20, 24, 28
      ~ August	1, 5, 9, 13, 17, 21, 25, 29
      ~ September	2, 6, 10, 14, 18, 22, 26, 30
      ~ October	4, 8, 12, 16, 20, 24, 28
      ~ November	1, 5, 9, 13, 17, 21, 25, 29
      ~ December	3, 7, 11, 15, 19, 23, 27, 31
  
#### Fast-GPS Control

    Maximum successful Fast-GPS attempts
      ~ 1 per hour; 4 per day
    Maximum failed Fast-GPS attempts
      ~ 3 per hour
    Overall maximum Fast-GPS attempts
      ~ 12 per day
    Supress Fast-GPS after good haulout location
      ~ enabled

### SPOT5 Programming Summary

#### Argos Transmissions
  
    Unused transmissions will be added to the next day's allowance.
    Maximum transmissions per day
      ~ 150
    Transmit on these hours
      ~ 1 - 4, 20 - 23
      ~ Tag will transmit during all of the hours before midnight on the first day of deployment
    Transmit on these days, using an absolute calendar
      ~ Jan: 1, 15, 
      ~ Feb: 1, 7, 13, 19, 25, 
      ~ Mar: 1, 7, 13, 19, 25, 31
      ~ Apr: 6, 12, 18, 24, 30
      ~ May: 6, 12, 18, 24, 30, 
      ~ Jun: 6, 12, 18, 24, 30
      ~ Jul: 6, 12, 18, 24, 30, 
      ~ Aug: 1, 15, 
      ~ Sep: 1, 15, 
      ~ Oct: 1, 15, 
      ~ Nov: 1, 15, 
      ~ Dec: 1, 15, 

    Time at temperature histograms **are not collected**
    Percent dry timelines **are being collected**


## Argos Least-Squared vs. Kalman Processing

In 2011, Argos changed their location estimation algorithm to include a Kalman
filter algorithm. This replaced their previous algorithm which relied on a 
least-squares process. The Kalman filter algorithm provides additional error
data which is critical for many movement models. When this change was made, 
all data for tags in this project were reprocessed back to January of 2008. In
most cases, for each location estimate in the least-squares dataset, there is a
corresponding Kalman filter estimate for that satellite pass.

The reprocessed data and the originally delivered least-squares data were
merged and reconciled by Wildlife Computers within their data portal. In this
process, locations were matched by _ptt_, _date-time_, _satellite_ and 
_pass duration_. _date-time_ and _pass duration_ were matched with fuzzy
logic (i.e. they were allowed to not match exactly). In all cases, the reprocessed
kalman locations were determined to be authoritative.

## Wildlife Computers Data Portal

Data included in this package were downloaded from the Wildlife Computers
Data Portal using the **wcUtils** package. After download, additional re-structuring
and processing of the data was also done with the **wcUtils** package. 

Those with collaborator permissions can access these deployments directly by searching for the _kotzeb09_ projectid.

# Data Components

The **kotzeb0912** package is a data package for distribution of core data 
products resulting from this study. There are 7 data products distributed with
this package:

1. `kotzeb0912_locs`
    - Argos location estimates for all deployments
2. `kotzeb0912_gps`
    - GPS location estimates from Fastloc solutions
2. `kotzeb0912_status`
    - tag status and performance messages for all deployments
3. `kotzeb0912_depths`
    - dive-depth histogram data; cleaned and tidy'd
4. `kotzeb0912_durations`
    - dive-duration histogram data; cleaned and tidy'd
5. `kotzeb0912_tad`
    - time-at-depth histogram data; cleaned and tidy'd
6. `kotzeb0912_timelines`
    - hourly percent dry (haul-out) data; cleaned and tidy'd
7. `kotzeb0912_deployments`
    - bearded seal morphometric and tag deployment data
    
## kotzeb0912_locs

```{r data-locs}
data("kotzeb0912_locs")
dplyr::glimpse(kotzeb0912_locs)
```

deployid
  ~ An alphanumeric string that uniquely identifies the
    deployment. 
    
    Since a deployment is a unqiue combination of animal and tag,
    the **deployid** is a concatenation of the animal id (**speno**) and
    the tag serial number (**serialnum**).
    
    By examining the **deployid**, one can discern several key details about
    the particular deployment. The first two characters in the **deployid**
    will specify the species of animal the tag was deployed on. These letters
    correspond to the genus and species (e.g. bearded seals (*Erignathus barbatus*)
    would be represented by *EB*). The next four characters represent the year
    the deployment started. The next section is separated by an underscore
    and is a unique number assigned to the animal. The first 11 characters in
    the **deployid** correspond to the **speno** for the deployment animal. The 
    next section (also separated by an underscore) corresponds to the serial
    number for the tag.
    
ptt
  ~ The Argos transmitter (PTT) identifier. 
    
    It is important to understand that a PTT identifier can be assigned to
    different tags over the years. The PTT identifier should not be considered a
    unique identifier.
    
instr
  ~ An alphanumeric string that identifies the make/model class of the tag
    
    ```{r unique-instr}
    unique(kotzeb0912_locs$instr)
    ```
    
    Mk10
      ~ Argos location tags with additional sensors for recording 
        depth/dive behavior, temperature, conductivity. For this project, 
        _Mk10_ tags were adhered to the hair of the seal on the head.
        
    SPOT
      ~ Argos location tags with conductivity. No pressure transducer for 
        recording depth or dive behavior. For this project, _SPOT_ tags were
        attached to the rear flipper.
        
date_time
  ~ A POSIXct value representing the UTC time for the location estimate.
    
    ```{r range-datetime}
    range(kotzeb0912_locs$date_time)
    ```
  
    Due to the nature of Argos data that is still not fully understood, there
    can be more than one record with the the same **date_time** value for the
    same deployment. See **unique_posix** for an adjusted value with no
    duplicate times.
    
type
  ~ An character string that specifies the location estimate is
    based on the Argos process
    
    ```{r unique-type}
    unique(kotzeb0912_locs$type)
    ```
 
 quality
  ~ An alphanumeric value cooresponding to the Argos location quality.
    
    ```{r unique_quality}
    unique(kotzeb0912_locs$quality)
    ```
  
    Argos location estimates were traditionally classified with a quality value
    that provides general guidelines regarding the error associated with the
    location estimates. From better to worse, the possible values are *3*, *2*,
    *1*,*0*,*A*,*B*, and *Z*. *Z* values should be removed from any analysis and
    are only included here for completeness.
    
    In recent years, Argos has provided better estimates
    of error using their Kalman filter algorithm. See **error_radius**,
    **error_semimajor_axis**, **error_semiminor_axis**, and
    **error_ellipse_orientation**.

latitude, longitude
  ~ Coordinates (decimal degrees, datum=WGS84) of the location estimate
  
error\_radius, error\_semimajor\_axis, error\_semiminor\_axis,error\_ellipse\_orientation
  ~ Error parameters provided as part of the Kalman filter algorithm
  
unique_posix
  ~ The same as **date_time** except duplicate values have been eliminated by
    increasing the value of one duplicate by 1 second.

## kotzeb0912_gps

```{r data-gps}
data("kotzeb0912_gps")
dplyr::glimpse(kotzeb0912_gps)
```

deployid
  ~ An alphanumeric string that uniquely identifies the
    deployment. 
    
    Since a deployment is a unqiue combination of animal and tag,
    the **deployid** is a concatenation of the animal id (**speno**) and
    the tag serial number (**serialnum**).
    
    By examining the **deployid**, one can discern several key details about
    the particular deployment. The first two characters in the **deployid**
    will specify the species of animal the tag was deployed on. These letters
    correspond to the genus and species (e.g. bearded seals (*Erignathus barbatus*)
    would be represented by *EB*). The next four characters represent the year
    the deployment started. The next section is separated by an underscore
    and is a unique number assigned to the animal. The first 11 characters in
    the **deployid** correspond to the **speno** for the deployment animal. The 
    next section (also separated by an underscore) corresponds to the serial
    number for the tag.

count
  ~  
  
time_offset
  ~  
  
locnumber
  ~  

hauled_out
  ~ a binary (1,0) indication whether the tag was in haul-out mode at the time the fastloc
  snapshot was captured

satellites
  ~ total number of satellites captured in the snapshot. substract the number of bad_sats to
    get the total number of viable satellites

initlat, initlon
  ~ Coordinates (decimal degrees, datum=WGS84) of the seed location used
  
inittime
  ~ A POSIXct value representing the UTC time for the seed location used
  
inittype
  ~ the source data for the seed coordinates: GPS, User or an letter corresponding to the Argos satellite

latitude, longitude
  ~ Coordinates (decimal degrees, datum=WGS84) of the location estimate

bad_sats
  ~ number of satellites with poor information in the snapshot and not usable
  
residual
  ~  

time_error
  ~  

date_time
  ~ A POSIXct value representing the UTC time for the fastloc location estimate.
    
    ```{r range-gps-datetime}
    range(kotzeb0912_gps$date_time)
    ```

## kotzeb0912_status

```{r data-status}
data("kotzeb0912_status")
dplyr::glimpse(kotzeb0912_status)
```

## kotzeb0912_depths

```{r data-depths}
data("kotzeb0912_depths")
dplyr::glimpse(kotzeb0912_depths)
```

deployid
  ~ An alphanumeric string that uniquely identifies the
    deployment. 
    
    Since a deployment is a unqiue combination of animal and tag,
    the **deployid** is a concatenation of the animal id (**speno**) and
    the tag serial number (**serialnum**).
    
    By examining the **deployid**, one can discern several key details about
    the particular deployment. The first two characters in the **deployid**
    will specify the species of animal the tag was deployed on. These letters
    correspond to the genus and species (e.g. bearded seals (*Erignathus barbatus*)
    would be represented by *EB*). The next four characters represent the year
    the deployment started. The next section is separated by an underscore
    and is a unique number assigned to the animal. The first 11 characters in
    the **deployid** correspond to the **speno** for the deployment animal. The 
    next section (also separated by an underscore) corresponds to the serial
    number for the tag.

datadatetime
  ~ A POSIXct value representing the UTC time for the start of the time bin.
  
    Dive behavior is summarized into user specified time bins. In this
    study, 6 hour bins were chosen.

num_dives
  ~ The number of dives with a max depth that occured during the corresponding
    time period and within the corresponding depth bin. This value is capped
    at 255. Thus, in cases when the actual number of dives exceeds 255 some 
    scaling occurs.
    
bin
  ~ The bin label in the original data files provided by Wildlife Computers

limits
  ~ The upper (numerically, not in the water column) limit of the depth bin in meters.
  
    The limits are determined from examining the data files provided from the
    Wildife Computers Data Portal. If the user has properly specified the 
    programming schema for this deployment, those values are extracted and
    included in the data frame for easy reference.

## kotzeb0912_durations

```{r data-durations}
data("kotzeb0912_durations")
dplyr::glimpse(kotzeb0912_durations)
```

deployid
  ~ An alphanumeric string that uniquely identifies the
    deployment. 
    
    Since a deployment is a unqiue combination of animal and tag,
    the **deployid** is a concatenation of the animal id (**speno**) and
    the tag serial number (**serialnum**).
    
    By examining the **deployid**, one can discern several key details about
    the particular deployment. The first two characters in the **deployid**
    will specify the species of animal the tag was deployed on. These letters
    correspond to the genus and species (e.g. bearded seals (*Erignathus barbatus*)
    would be represented by *EB*). The next four characters represent the year
    the deployment started. The next section is separated by an underscore
    and is a unique number assigned to the animal. The first 11 characters in
    the **deployid** correspond to the **speno** for the deployment animal. The 
    next section (also separated by an underscore) corresponds to the serial
    number for the tag.

datadatetime
  ~ A POSIXct value representing the UTC time for the start of the time bin.
  
    Dive behavior is summarized into user specified time bins. In this
    study, 6 hour bins were chosen.
    
num_dives
  ~ The number of dives with a duration (seconds) that occured during the corresponding
    time period and within the corresponding duration bin. This value is capped
    at 255. Thus, in cases when the actual number of dives exceeds 255 some 
    scaling occurs. 

bin
  ~ The bin label in the original data files provided by Wildlife Computers

limits
  ~ The upper limit of the duration bin in seconds.
  
    The limits are determined from examining the data files provided from the
    Wildife Computers Data Portal. If the user has properly specified the 
    programming schema for this deployment, those values are extracted and
    included in the data frame for easy reference.

## kotzeb0912_tad

```{r data-tad}
data("kotzeb0912_tad")
dplyr::glimpse(kotzeb0912_tad)
```

deployid
  ~ An alphanumeric string that uniquely identifies the
    deployment. 
    
    Since a deployment is a unqiue combination of animal and tag,
    the **deployid** is a concatenation of the animal id (**speno**) and
    the tag serial number (**serialnum**).
    
    By examining the **deployid**, one can discern several key details about
    the particular deployment. The first two characters in the **deployid**
    will specify the species of animal the tag was deployed on. These letters
    correspond to the genus and species (e.g. bearded seals (*Erignathus barbatus*)
    would be represented by *EB*). The next four characters represent the year
    the deployment started. The next section is separated by an underscore
    and is a unique number assigned to the animal. The first 11 characters in
    the **deployid** correspond to the **speno** for the deployment animal. The 
    next section (also separated by an underscore) corresponds to the serial
    number for the tag.

datadatetime
  ~ A POSIXct value representing the UTC time for the start of the time bin.
  
    Dive behavior is summarized into user specified time bins. In this
    study, 6 hour bins were chosen.

pct_tad
  ~ Percent time-at-depth represents the percentage of time the tag
    spent within the corresponding depth bin. Values across the depth bins for
    a given 6 hour time period should add to approximately 100.

bin
  ~ The bin label in the original data files provided by Wildlife Computers

limits
  ~ The upper (numerically, not in the water column) limit of the dive depth 
    bin in meters
  
    The limits are determined from examining the data files provided from the
    Wildife Computers Data Portal. If the user has properly specified the 
    programming schema for this deployment, those values are extracted and
    included in the data frame for easy reference.

## kotzeb0912_timelines

```{r data-timelines}
data("kotzeb0912_timelines")
dplyr::glimpse(kotzeb0912_timelines)
```

deployid
  ~ An alphanumeric string that uniquely identifies the
    deployment. 
    
    Since a deployment is a unqiue combination of animal and tag,
    the **deployid** is a concatenation of the animal id (**speno**) and
    the tag serial number (**serialnum**).
    
    By examining the **deployid**, one can discern several key details about
    the particular deployment. The first two characters in the **deployid**
    will specify the species of animal the tag was deployed on. These letters
    correspond to the genus and species (e.g. bearded seals (*Erignathus barbatus*)
    would be represented by *EB*). The next four characters represent the year
    the deployment started. The next section is separated by an underscore
    and is a unique number assigned to the animal. The first 11 characters in
    the **deployid** correspond to the **speno** for the deployment animal. The 
    next section (also separated by an underscore) corresponds to the serial
    number for the tag.

datadatetime
  ~ A POSIXct value representing the UTC time for the start of the hour bin.
  
    Percent timeline data is summarized into hourly time bins.

percent_dry
  ~ A percentage of the given hour the tag was dry (out of the water)
  
    Possible values include: *0*,*3*,*5,*,*10*,*20*,*30*,*40*,*50*,*60*,*70*,*80*,
    *90*,*95*,*97*, and *100*

## kotzeb0912_deployments

```{r data-deployments}
data("kotzeb0912_deployments")
dplyr::glimpse(kotzeb0912_deployments)
```

# Publications and Data Availability

The final contract report for this research is available from the [Bureau of
Ocean and Energy Management](http://www.boem.gov/BOEM-Newsroom/Library/Publications/2013/BOEM_2013_001150_pdf.aspx)

This R data package and associated vignette documents should be cited as

> London, Josh M., M.F. Cameron, and P.L. Boveng. 2015. kotzeb0912: An R package
of data from a satellite telemetry study of bearded seals in Alaska.

# Acknowledgments and Funding

The research described here and the included data were obtained with significant
financial contributions from the U.S. Department of Interior's Bureau of Ocean
and Energy Management (BOEM) [^4] and the U.S. Department of Commerce's National
Oceanic and Atmospheric Administration.

In addition to funding, significant leadership, participation and expertise was
provided by the Kotzebue IRA and members of the Kotzebue community. 

[^4]: funding administered under the Inter-agency Agreement M07RG13317

